An Empirical Study of Invasion and Resistance for Iterated Prisoner's Dilemma Strategies

### 2017-05-11 main.tex
Add training section.

To carry out this large numerical experiment, 164 strategies are used from
\cite{axelrodproject}. These include 161 default strategies in the library at
the time (excluding strategies classified as having a long run time and those
that make use of the length of the game) as well as
the following 6 finite state machine machine strategies \cite{Ashlock2006}:

% TODO Include details about the 3/6 extra FSM strategies.
% Possibly draw pictorial representations?
% Remove the FSM strategies if they don't perform well.
% TODO Decide if we include all 6 (so the same original 3 with the wrong initial
% state) in the entire analysis.

Appendix~\ref{app:list_of_players} shows all the players in question. More
information about each player can be obtained in the documentation for
\cite{axelrodproject}. There are \input{./tex/num_stochastic.tex}stochastic and
\input{./tex/num_deterministic.tex}deterministic strategies. Their memory
depth, defined by the number of rounds of history used by the strategy each round,
is shown in Table~\ref{tbl:memory_depth_count}. The memory depth is infinite if
the strategy uses the entire history of play (whatever its length). For example,
a strategy that utilizes a handshaking mechanism where the opponents actions
on the first few rounds of play determines the strategies subsequent behavior
would have infinite memory depth.

>>>>

To carry out this large numerical experiment, 164 strategies are used from the
axelrod library: \cite{axelrodproject}.

The axelrod library contains many machine learning based strategies trained
with reinforcement learning algorithms. This work includes three additional
strategies trained specifically to excel at the Moran process.
Appendix~\ref{app:list_of_players} shows all the players in question. More
information about each player can be obtained in the documentation for
\cite{axelrodproject}. There are \input{./tex/num_stochastic.tex}stochastic and
\input{./tex/num_deterministic.tex}deterministic strategies. Their memory
depth, defined by the number of rounds of history used by the strategy each round,
is shown in Table~\ref{tbl:memory_depth_count}. The memory depth is infinite if
the strategy uses the entire history of play (whatever its length). For example,
a strategy that utilizes a handshaking mechanism where the opponents actions
on the first few rounds of play determines the strategies subsequent behavior
would have infinite memory depth.

All of the
strategies are trained with an evolutionary algorithm that perturbs strategy
% TODO Add a generic text book reference
parameters and optimizes an objective function, such as the mean total score
against all other opponents or the mean fixation probability over a large
number of Moran processes. The objective functions can include standard
variations such as noisy matches, varying the number of rounds of play, and the
other typical parameters used for the IPD. Variation is introduced via mutation
and crossover of parameters, and the best performing strategies are carried to
the next generation along with new variants. Similar methods appear in the
literature \cite{Ashlock2006}.

All of our training code is available in the
Axelrod repository with documentation to train new strategies easily. Training
% TODO Include link etc
typically takes less than 200 iterations and can be completed within several
hours on commodity hardware.  The three strategies trained specifically for
this study are Trained FSM 1, 2, and 3 (TF1 - TF3), based on finite state
machines of 16, 16, and 8 states respectively (see Figures \ref{}).
% TODO Include images of FSMs
These strategies were trained with the objective function of mean fixation
probabilities for Moran processes starting at initial population states
consisting of \(N/2\) individuals of the training candidates and \(N/2\)
individuals of an opponent strategy, for all of the short run time strategies
in the library (approximately 150 opponents at the time of training):

\begin{itemize}
	\item TF1 \(N=12\), 0\% noise, 10000 repetitions per match
	\item TF2 \(N=10\), 0\% noise, 10000 repetitions per match
	\item TF3 \(N=8\), 1\% noise, 100 repetitions per match
\end{itemize}

Each matchup of players was run to fixation for the specified number of
repetitions to estimate the absorption probabilities. The trained algorithms
were run for less than 50 generations.

TF1 has an initial handshake of CCD and cooperates if the opponent matches.
However if the opponent later defects, TF1 will respond in kind, so the
handshake is not permanent. Only one player (Prober 4 \cite{prison}) manages to
achieve cooperation with TF1 after about 20 rounds of play.

TF2 always starts
with CD and will defect against opponents that start with DD. It plays CDD
against itself and then cooperates thereafter. There is a longer complex
handshake which eventually results in mutual cooperation with Firm but Fair,
Fortress3, Fortress4, and Grofman (always) and Evolved HMM 5 and GTFT
(depending on the random seed).

TF3 cooperates and defects with
various cycles depending on the opponent's actions. TF3 will mutually
cooperate with any strategy and only tolerates a few defections before
defecting for the rest of match. It is similar to but not exactly the same as
Fool Me Once, a strategy that cooperates until the opponent has defected twice
(not necessarily consecutively), and defects indefinitely thereafter. Though a
product of training with a Moran objective, it was selected for inclusion
because of the lack of an apparent handshake.

For both TF1 and TF2 a handshake
mechanism naturally emerges from the structure of the underlying finite state
machine. This behavior is an outcome of the evolutionary process and is in no
way hard-coded or included via an additional mechanism. Some of the existing
strategies in the library were trained on earlier versions of the library
(smaller total opponents used in training) based on neural networks, lookup
tables and stochastic variants trained with a particle swarm algorithm, finite
state machines and a stochastic variant mimicking hidden Markov models, and
neural networks. These strategies were trained to win IPD tournaments using an
objective function that computes the mean match payoffs against a collection of
opponents. Detailed descriptions and the performance of these strategies in IPD
tournaments will be described in another manuscript.
% TODO Think of how to reference this upcoming manuscript
% TODO Move the section below to the discussion for particular
% cases and/or conclusion
These strategies trained with the mean score
objective are among the best invaders in the library but are not as resistant
to invasion as the strategies trained using a Moran objective function (see
Tables \ref{}). These strategies include trained finite state machine
strategies, but they do not appear to have handshaking mechanisms. Therefore it
is reasonable to conclude that the objective function is the cause of the
emergence of handshaking mechanisms.
The payoff maximizing strategies
typically will not defect before the opponent's first defection, possibly
because the training strategy collection contains a significant portion of
strategies such as Grudger and Fool Me Once that retaliate harshly by defecting
for the remainder of the match if the opponent has more than a small number of
cumulative defections. Paradoxically it is necessary to defect in order to
achieve mutual cooperation with opponents using the same strategy but not with
other opponents. A handshake requires at least one defection and there is
selective pressure to defect as few times as possible to achieve the
self-recognition mechanism. It is also unwise to defect on the first move as
some strategies additionally retaliate first round defections. So the
handshakes used by TF1 and TF2, and CS, are in some sense optimal.  These
discoveries may have significant ramifications regarding the evolution of
cooperation and forgiveness in biological organisms and social interactions
between humans.
